---
title: "Single-Cell RNA-seq Analysis of Human PBMCs with Bioconductor- sample integration"
author: "Adrien Jolly"

---

---

# Load Required Packages

```{r setup, message=FALSE, warning=FALSE}
library(DropletUtils)
library(AnnotationHub)
library(SingleCellExperiment)
library(scater)
library(scran)
library(bluster)
library(batchelor)
library(ggplot2)
library(patchwork)
library(SingleR)
library(celldex)
library(DESeq2)
library(edgeR)
library(scDblFinder)
```

---

# Load 10x Data

3 samples


```{r}
# loading the samples
sce1 <- readRDS("~/Nextcloud2/TutorialSingleCellRNAseq/tutorial2602/sce1.rds")

sce2 <- readRDS("~/Nextcloud2/TutorialSingleCellRNAseq/tutorial2602/sce2.rds")

sce3 <- readRDS("~/Nextcloud2/TutorialSingleCellRNAseq/tutorial2602/sce3.rds")

saveRDS(sce,"scecombined2702.rds")

sce=readRDS("scecombined2702.rds")

```

# now we combine the datasets
```{r}
ctsce1= sce1$celltype
ctsce2= sce2$celltype
ctsce3= sce3$celltype


colData(sce1)=colData(sce1)[1:2]
colData(sce2)=colData(sce2)[1:2]
colData(sce3)=colData(sce3)[1:2]

sce1$celltype = ctsce1
sce2$celltype = ctsce2
sce3$celltype = ctsce3


# keep only common genes
common_genes <- Reduce(intersect,list(rownames(sce1),rownames(sce2),rownames(sce3)))
                            
                            
sce1 <- sce1[common_genes, ]
sce2 <- sce2[common_genes, ]
sce3 <- sce3[common_genes, ]

#if necessary, we can remove some columns which do not match between samples

#colData(sce1)=colData(sce1)[1:2]
#colData(sce2)=colData(sce2)[1:2]
#colData(sce3)=colData(sce3)[1:2]

# exclude dimensional reduction which we will generate again

reducedDim(sce1,"PCA") <- NULL
reducedDim(sce2,"PCA") <- NULL
reducedDim(sce3,"PCA") <- NULL

reducedDim(sce1,"UMAP") <- NULL
reducedDim(sce2,"UMAP") <- NULL
reducedDim(sce3,"UMAP") <- NULL

#rescale everything 

sce.list <- multiBatchNorm(sce1, sce2, sce3)

#add information for each donor
sce.list[[1]]$donor <- "D1"
sce.list[[2]]$donor <- "D2"
sce.list[[3]]$donor <- "D3"


#check that cell annotation columns are the same
setdiff(colnames(colData(sce1)),colnames(colData(sce2)))

sce <- do.call(cbind, sce.list)

table(sce$donor)
```

# model variance fo combined dataset
```{r}
dec <- modelGeneVar(sce)
fit.default <- metadata(dec)
plot(dec$mean,
     dec$total,
     pch=16,
     cex=0.5,
     xlab="Mean expression",
     ylab="Total variance",
     main="Mean-Variance Plot")
curve(fit.default$trend(x), col="dodgerblue", add=TRUE, lwd=2)
```

# get top Highly variable Genes

```{r}
hvg <- getTopHVGs(dec, n=2000)

head(hvg)

```

# PCA

Dimensionality reduction on selected features.

```{r}
sce <- runPCA(sce, subset_row=hvg)

```

For further analysis, we want to select a number of components which are informative and discard non informative PCs

```{r}
var_explained <- attr(reducedDim(sce, "PCA"), "percentVar")

plot(var_explained,
     type="b",
     xlab="PC",
     ylab="Percent variance explained",
     main="Elbow Plot")
```

#run UMAP
```{r}

set.seed(1001010) # initialize UMAP (for reproducibility)
sce <- runUMAP(sce, dimred="PCA", n_dimred=7)
```



#visualize based on manual annotation of single samples

```{r}
plotUMAP(sce, colour_by="donor",point_size=0.3)
```



#marker genes per cluster identification
```{r}
markers <- findMarkers(sce, sce$clustersCorrected)


FDRclust6=markers[[6]]$FDR

length(which(FDRclust6<0.05))



markersClust2=markers[[2]]

#markers for a given cluster will have high summary log2FC

topClust2=markersClust2[order(markersClust2$summary.logFC,decreasing=TRUE),]



rownames(topClust2)[1:30]




```


# batch correction:

in order to perform a single coherent clustering we may need to create a corrected dimensional reduction which eliminate the sample effect. 

```{r}

#we batch correct using fastMNN
sce.mnn <- fastMNN(sce.list,d = 6)
reducedDim(sce.mnn, "corrected")
reducedDim(sce, "MNN") <- reducedDim(sce.mnn, "corrected")

# we perform a UMAP using the corrected dimensionality reduction and check whether the donor effect is corrected.
sce <- runUMAP(
    sce,
    dimred = "MNN"
)

plotUMAP(sce, colour_by = "donor",point_size=0.5)

```



#  Clustering and marker detection
 
 we now recluster based on MNN corrected factors

```{r}


clustersCorrected <- clusterRows(reducedDim(sce, "MNN")[,1:6], NNGraphParam(cluster.fun="louvain",cluster.args=list(resolution=0.2)))
sce$clustersCorrected <- factor(clustersCorrected)

```

#Visualize corrected clusters

```{r}
plotUMAP(sce, colour_by="clustersCorrected")
```

#check the proportion of cells of each patient per cluster

```{r}

# table of cells per patient, and per cluster
CLusterPerDonor <- table(sce$clustersCorrected,sce$donor)


# % of cells of each patient per cluster
props <- prop.table(CLusterPerDonor, margin = 1)

#plotting the proportion with a heatmap
library(pheatmap)

pheatmap(props,cluster_rows = FALSE, cluster_cols = FALSE,ylim=c(0,1))

```

#  Create Pseudobulk Counts

cells are not biological replicates, statistical tools to study differential gene expression between populations

```{r}
table(sce$donor)
```

#Aggregate by donor and new clusters (clustersCorrected) (generated from combined dataset) to create a single cell experiment with nrow= n genes and ncol = (n donors * m clusters). 

```{r}

#The function sums raw cell counts within groups specified by the user
pseudo <- aggregateAcrossCells(
  sce,
  ids=DataFrame(donor=sce$donor,
                cluster=sce$clustersCorrected)
)
```

---

#. Differential Expression with DESeq2

#We perform differential gene expression between clusters using DESeq. A tool using generalized linear models to test whether a specific factor (like cluster, batch effect..) affect the expression of given genes.

```{r}


# we create the DESeq2 object with the raw count matrix (DESeq handles normalization internally), and colData (annotations at the level of the pseudobulk). the "design" option specifies the linear model, in this case we set the set out that gene expression is only affected by the cluster (clustersCorrected). 
dds <- DESeqDataSetFromMatrix(
  countData=assay(pseudo, "counts"),
  colData=colData(pseudo),
  design=~clustersCorrected
)

#keep exclusively column of interest (since the column were generated from the column of the single cell experiments most of them are now meaningless at bulk level)
colData(dds) <- colData(dds)[6]



dds <- DESeq(dds)
#compare expression in cluster 1 vs cluster 2.
res <- results(dds,contrast=c("clustersCorrected","1","2"))

```

generate volcano plot:

```{r}

DifExpr <- as.matrix(res[order(res$padj),])

DifExpr <- DifExpr[which(DifExpr[,"padj"]<0.05),]

markers <- rownames(DifExpr[which(abs(DifExpr[,"log2FoldChange"])>=5&DifExpr[,"padj"]<=0.0001),])


p <- EnhancedVolcano(res, 
                rownames(res),
                x ="log2FoldChange", 
                y ="padj",
                selectLab=markers,
                max.overlaps = Inf,
                xlab = bquote(~Log[2]~ 'fold change'),
                pCutoff = 0.05,
                FCcutoff = 0.5,
                pointSize = 1.0,
                labSize = 4.0,
                labCol = 'black',
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 0.75,
                colConnectors = 'black',
                title="cluster x vs cluster y - comp",
                subtitle="FDR<=0.05, Absolute FC>=0.5",
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                xlim=c(-3,5)
)

p
ggsave("EnhancedVolcano.pdf", plot = p, width = 8, height = 6)

```
